MWA tied-array processing III: Microsecond time resolution via a polyphase synthesis filter

### 2020-05-09 ipfb.tex
Wrote up how to find the least squares solution, and also made the SNR plot pretty

\section{Synthesis filter}
\label{sec:ipfb}

@@

Back-to-back analysis-synthesis filters can be designed so that the original time series can be perfectly reconstructed without any loss of signal (i.e. $\hat{x}[n] = x[n]$ exactly), and the system as a whole can be thought of as an identity operation.
The condition for perfect reconstruction can be found by substituting the analysed spectrum obtained from Eq. \eqref{eqn:wdft} into the synthesis operation defined in Eq. \eqref{eqn:ipfb}.
It can be shown that, for a critically sampled system,
\begin{equation}
    \begin{aligned}
        \hat{x}[n]
            &= \sum_{s = -\infty}^{\infty} x[n - sM]
               \sum_{m = -\infty}^{\infty} f_m[n]\, h_m[sM - n] \\
            &= x[n] \sum_{m = -\infty}^{\infty} f_m[n]\, h_m[-n] +{} \\
            &\quad \sum_{\substack{s = -\infty \\ s \ne 0}}^{\infty} x[n - sM]
               \sum_{m = -\infty}^{\infty} f_m[n]\, h_m[sM - n]
    \end{aligned}
    \label{eqn:backtoback}
\end{equation}
where, following \citet{Crochiere1983}, we have adopted the shorthand notation
\begin{equation}
    \begin{aligned}
        h_m[n] &\equiv h[n - mM], \\
        f_m[n] &\equiv f[n - mM].
    \end{aligned}
\end{equation}
It can be seen that perfect reconstruction corresponds to the case when the only non-zero contribution comes from the $s = 0$ term, giving rise to the necessary and sufficient condition
\begin{equation}
    \sum_{m = -\infty}^{\infty} f[n-mM]\,h[(m - s)M - n] = \Kron
    \label{eqn:invertibility_condition}
\end{equation}
for all values of $n$, where $\Kron$ is the Kronecker delta.

@@

In the context of an astrophysical signal, it is desirable to quantify the reconstruction error in terms of the signal-to-noise ratio (S/N).
If we assume that individual samples follow, and are dominated by, the same Gaussian noise statistics, then the reduction in S/N of a reconstructed sample can be estimated by considering the fraction of the power of the reconstructed sample that came from the original sample:
\begin{equation}
    \frac{[\text{S/N}]_\text{recon}}{[\text{S/N}]_\text{orig}}
      = \frac{\left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_m[-n]\right)^2}{
            \sum\limits_{s = -\infty}^{\infty}
            \left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_m[sM - n]\right)^2
        }
\end{equation}
This expression is invariant under the substitution $n \rightarrow n + \lambda M$, $\lambda \in \mathbb{Z}$, which indicates that the (average) reconstruction error is only a function of where the sample in question falls within a tap.
Consequently, any synthesis filter (except the exact inverse of the analysis filter, if it exists) will introduce a ``ringing'' effect into the reconstructed time series that has a period equal to the size of the PFB tap.

\subsection{Design constraints}

Eq. \eqref{eqn:backtoback} suggests that a reconstructed time sample potentially contains contributions from up to $P$ original samples, spaced at intervals of one tap.

Whether this condition is perfectly met clearly depends on the choice of analysis and synthesis filters.
However, even a synthesis filter which only approximately satisfies Eq. \eqref{eqn:invertibility_condition} for a given analysis filter (i.e. $\lambda(s) \approx \Kron$) will produce a time series that approximates the original coarse channel time series (i.e. $\hat{x}[n] \approx x[n]$).
In practice, the choice of synthesis filter is often obliged to be a compromise between ease of implementation and an acceptable level of imperfection in the reconstructed time series.

For the analysis filter $h[n]$ defined in Eq. \eqref{eqn:filter}, it is found that choosing the synthesis filter
\begin{equation}
    f[n] = \frac{h[-n]}{H}
    \label{eqn:synthesisfilter}
\end{equation}
where $H = \sum_{n=0}^{N-1} h[n]$ is a normalising factor, meets the invertibility condition sufficiently closely for our purposes, as demonstrated below.
With the analysis and synthesis filters thus defined, the value of Eq. \eqref{eqn:invertibility_condition} can be calculated and is illustrated in Fig. \ref{fig:inverse_condition}.
It is readily seen that $\lambda(s)$ closely approximates, but does not exactly equal, $\Kron$.
The quantitative effect of this analysis-synthesis pair is discussed in Section \ref{sec:fidelity}.

%\begin{figure}
%    \centering
%    \includegraphics[width=\columnwidth]{inverse_condition.eps}
%    \caption{The condition for invertibility, Eq. \eqref{eqn:invertibility_condition}, demonstrated for the analysis filter given in Eq. \eqref{eqn:filter} and the synthesis filter given in Eq. \eqref{eqn:synthesisfilter}. The fact that $\lambda(s)$ only approximates the delta function gives rise to spurious artefacts (``temporal echoes'') in the time domain, which can be seen in Fig. \ref{fig:impulse_response}.}
%    \label{fig:inverse_condition}
%\end{figure}

@@

\subsection{Implementation}

>>>>

\section{Synthesis filters}
\label{sec:ipfb}

@@

Back-to-back analysis-synthesis filters can be designed so that the original time series can be perfectly reconstructed without any loss of signal (i.e. $\hat{x}[n] = x[n]$ exactly), and the system as a whole can be thought of as an identity operation.
The condition for perfect reconstruction can be found by substituting the analysed spectrum obtained from Eq. \eqref{eqn:wdft} into the synthesis operation defined in Eq. \eqref{eqn:ipfb}.
For a critically sampled system,
\begin{equation}
    \begin{aligned}
        \hat{x}[n]
            &= \sum_{s = -\infty}^{\infty} x[n - sM]
               \sum_{m = -\infty}^{\infty} f_m[n]\, h_{m-s}[-n] \\
            &= x[n] \sum_{m = -\infty}^{\infty} f_m[n]\, h_m[-n] +{} \\
            &\quad \sum_{\substack{s = -\infty \\ s \ne 0}}^{\infty} x[n - sM]
               \sum_{m = -\infty}^{\infty} f_m[n]\, h_{m-s}[-n]
    \end{aligned}
    \label{eqn:backtoback}
\end{equation}
where, following \citet{Crochiere1983}, we have adopted the shorthand notation
\begin{equation}
    \begin{aligned}
        h_\lambda[n] &\equiv h[n - \lambda M], \\
        f_\lambda[n] &\equiv f[n - \lambda M].
    \end{aligned}
    \notag
\end{equation}
It can be seen that perfect reconstruction corresponds to the case when the only non-zero contribution comes from the $s = 0$ term, giving rise to the necessary and sufficient condition
\begin{equation}
    \sum_{m = -\infty}^{\infty} f_m[n]\,h_{m-s}[-n] = \Kron
    \label{eqn:invertibility_condition}
\end{equation}
for all values of $n$, where $\Kron$ is the Kronecker delta.

@@

In the context of an astrophysical signal, it is desirable to quantify the reconstruction error in terms of the signal-to-noise ratio (S/N).
If we assume that individual samples follow, and are dominated by, the same Gaussian noise statistics, then the reduction in S/N of a reconstructed sample can be estimated by considering the fraction of the power of the reconstructed sample that came from the original sample:
\begin{equation}
    \frac{[\text{S/N}]_\text{recon}}{[\text{S/N}]_\text{orig}}
      = \frac{\left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_m[-n]\right)^2}{
            \sum\limits_{s = -\infty}^{\infty}
            \left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_{m-s}[-n]\right)^2
        }.
\end{equation}
This expression is invariant under the substitution $n \rightarrow n + \lambda M$, $\lambda \in \mathbb{Z}$, which indicates that the (average) reconstruction error is only a function of where the sample in question falls within a tap.
Consequently, any synthesis filter (except the exact inverse of the analysis filter, if it exists) will introduce a ``ringing'' effect into the reconstructed time series that has a period equal to the size of the PFB tap.

\subsection{Optimal and sub-optimal filter designs}

Minimising the loss of S/N is equivalent to solving Eq. \eqref{eqn:invertibility_condition} in a least squares sense.
Since any given reconstructed sample only receives contributions from samples spaced one tap apart, Eq. \eqref{eqn:invertibility_condition} can be thought of as $M$ independent conditions, one for each tap position, $n = 0, 1, 2, \dots, M-1$.
Thus, considering each tap position separately, each condition can be expressed as a minimal matrix equation,
\begin{equation}
    H^{(n)}F^{(n)} = D,
\end{equation}
where $H^{(n)}$, $F^{(n)}$, and $D$ are matrices whose elements are given by
\begin{equation}
    \begin{aligned}
        H^{(n)}_{ij} &= h_{P-1+j-i}[-n], \\
        F^{(n)}_j    &= f_j[n], \\
        D_i          &= \delta^{P-1}_i,
    \end{aligned}
\end{equation}
where $P$ is the number of taps in the analysis filter, and the size of $F^{(n)}$ is set to the desired number of (non-zero) taps in the synthesis filter, $P^\prime$.
The indices are chosen such that, owing to the finite size of the analysis filter, the smallest $H^{(n)}$ that captures every non-trivial term in Eq. \eqref{eqn:invertibility_condition} is the $(P^\prime+P-1) \times P^\prime$ matrix
\begin{equation}
    H^{(n)} =
    \begin{bmatrix}
       h_{P-1}[-n] & 0 & \cdots & 0 \\
       h_{P-2}[-n] & h_{P-1}[-n] & \cdots & 0 \\
       \vdots & & \ddots & \vdots \\
       h_0[-n] & h_1[-n] & \cdots & h_{P-1}[-n] \\
       \vdots & & \ddots & \vdots \\
       0 & \cdots & h_0[-n] & h_1[-n] \\
       0 & \cdots & 0 & h_0[-n]
    \end{bmatrix},
\end{equation}
with row number $i = P - 1$ (in $H^{(n)}$ and $D$) corresponding to the $s = 0$ term.

Once the matrices $H^{(n)}$, $F^{(n)}$, and $D$ have been defined, solution by least squares regression can proceed in the usual way, yielding the best-fit filter coefficients
\begin{equation}
    \hat{F}^{(n)} = \left({H^{(n)}}^T H^{(n)}\right)^{-1} {H^{(n)}}^T D,
\end{equation}
where ${H^{(n)}}^T$ is the transpose of $H^{(n)}$.
With this notation, the reconstruction error can be more simply expressed
\begin{equation}
    \frac{[\text{S/N}]_\text{recon}}{[\text{S/N}]_\text{orig}}
      = \frac{(\hat{D}_{P-1})^2}{\hat{D}^T \hat{D}},
    \label{eqn:snr_short}
\end{equation}
where $\hat{D} = H^{(n)} \hat{F}^{(n)}$.

Fig. \ref{fig:synthfilters} shows the solutions found for the MWA's analysis filter defined in Eq. \eqref{eqn:filter}, for 12-, 18-, and 24-tap synthesis filter sizes.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{example-image-a}
    \caption{Least-squares solutions for different filter sizes \dots}
    \label{fig:synthfilters}
\end{figure}
Despite the complexity of the synthesis filters, they all share the same basic form as the analysis filter.
This suggests that choosing a synthesis filter that is the mirror image of the analysis filter (i.e. $f[n] = h[-n]$), might also yield a reasonably good reconstruction, without having to calculate the optimal solutions numerically.
The performance of both the mirror filter and the set of optimal solutions (for different choices of $P^\prime$) can be evaluated straightforwardly using Eq. \eqref{eqn:snr_short}.
However, before comparing the results for various filters, there is another source of reconstruction error, specific to the MWA implementation of the PFB, that must first be considered.

\subsection{Quantisation error}

\subsection{Comparison of various filters}

@@

\subsection{MWA Implementation}
